---
title: "Vancouver street trees analysis with Tidyverse"
author: Mboya Daisy
date: 04 October 2024
format: html
editor: visual
---

#### 1. Introduction

This report presents an analysis of Vancouver's street trees using the `tidyverse` collection of R packages. The primary objectives include data cleaning, transformation, visualization, and statistical analysis using `dplyr`, `tidyr`, and `ggplot2`.

##### Learning Objectives:

-   Familiarization with the tidyverse environment
-   Data cleaning and basic analysis
-   Plotting and mapping aesthetics using `ggplot2`

#### 2. Dataset Description

The dataset is sourced from Vancouver City Open Data Portal and contains information on trees located on public streets. The key attributes include:

-   Tree ID
-   Tree species (Latin and common name)
-   Diameter at breast height (in inches)
-   Height class (integer scale from 0-10)
-   Street and neighborhood name

#### 3. Required Software and Libraries

```{r}
#| warning: false

# Load necessary packages 
library(tidyverse) 
library(dplyr) 
library(janitor) 
library(ggplot2) 
library(ggrepel)
```

#### 4. Data Cleaning and Transformation

##### 4.1. Checking and Reshaping `trees_count.csv`

```{r}
# Load the data 
trees_count <- data.frame(read_csv("trees_count.csv", show_col_types = FALSE))  
# Reshape the data to a tidy format 
trees_count <- pivot_longer(data = trees_count, cols = 2:ncol(trees_count),
                            names_to = 'neighbourhoods',values_to = 'count')
```

##### 4.2. Checking and Reshaping `trees_height_diameter.csv`

```{r}
# Load the data 
trees_h_d <- data.frame(read_csv("trees_height_diam.csv", show_col_types = FALSE))
# Reshape to wide format 
trees_h_d <- pivot_wider(data = trees_h_d, names_from = 'attribute', values_from = 'value')
```

##### 4.3. Cleaning `street_trees.csv` Column Names

```{r}
# Load the data 
street_trees <- data.frame(read_csv("street_trees.csv", 
                                    show_col_types = FALSE))  
# Rename columns for consistency 
street_trees <- dplyr::rename(street_trees,
                              tree_id = Tree.ID,
                              street = Street.Name,
                              neighbourhood = Neighbourh,
                              species = SpeciesName,
                              common_name = CommonName,
                              height_rank = hrank,
                              diameter_in = Diameter,
                              year_planted = YearPlanted)
```

##### 4.4. Handling Data Types and Unit Conversion

```{r}
# Convert diameter from inches to cm 
street_trees <- mutate(street_trees, diameter_cm = diameter_in * 2.54)  
# Convert height rank to ordered factor 
height_rank_unique <- unique(street_trees$height_rank) 

height_rank_ordered <- sort(height_rank_unique) 

street_trees$height_rank <- factor(street_trees$height_rank,
                                   levels = height_rank_ordered, ordered = TRUE)
```

##### 4.5. Removing Unrealistic Tree Diameters

```{r}
# Identify trees with unrealistic diameters 
num_zero_diameter <- nrow(filter(street_trees, diameter_cm == 0)) 

num_large_diameter <- nrow(filter(street_trees, diameter_cm > 300))  
# Filter to remove unrealistic values 
trees_clean <- filter(street_trees, diameter_cm != 0 & diameter_cm <= 300)
```

#### 5. Data Summarization

##### 5.1. Overall Summary of Tree Diameters

```{r}
# Compute summary statistics 
trees_summary_all <- summarize(trees_clean,diameter_mean = mean(diameter_cm),
                               diameter_min = min(diameter_cm),
                               diameter_max = max(diameter_cm),
                               diameter_sd = sd(diameter_cm),n_obs = n())

trees_summary_all
```

##### 5.2. Tree Summary by Species

```{r}
# Grouped summary by species 
trees_summary_sp <- trees_clean %>% 
  group_by(common_name) %>% 
  summarize(diameter_mean = mean(diameter_cm),
            diameter_min = min(diameter_cm),
            diameter_max = max(diameter_cm),
            diameter_sd = sd(diameter_cm),
            n_obs = n())

trees_summary_sp
```

##### 5.3. Top 5 Most Common Species

```{r}
# Identify the top 5 species 
top_5_sp <- trees_clean %>% 
  count(common_name) %>% 
  arrange(desc(n)) %>% 
  slice_head(n = 5)

top_5_sp
```

#### 6. Discussion and Insights

1.  The most common tree species in Vancouver is the Kwanzan Flowering Cherry, followed by Pissard Plum.
2.  South Cambie has the highest tree density, but it covers a small area, making it a local hotspot for street trees.
3.  Neighborhoods with high tree densities include Renfrew-Collingwood, Hastings-Sunrise, and Dunbar-Southlands, each having over 100 trees per km of street.
4.  Downtown has the smallest tree diameters on average, likely due to urban constraints and maintenance practices.
5.  Some species, such as Pyramidal European Hornbeam, tend to have smaller tree diameters, while Norway Maple has trees with a broader range of diameters.

#### 7. Conclusion

This analysis demonstrates how `tidyverse` can effectively process and analyze urban forestry data. Key findings include:

-   Tree diversity and distribution: Vancouver's street trees are varied, with some neighborhoods having a higher density than others.
-   Data cleaning is crucial: Identifying and removing incorrect data (such as zero-diameter trees) is necessary for accurate analysis.
-   Urban planning implications: Understanding tree distribution helps in urban greening efforts and policy decisions.

Further analysis could involve spatial mapping of tree distributions, investigating the impact of tree age on diameter, and evaluating environmental benefits such as carbon sequestration and shade provision.

#### 8. References

-   Vancouver Open Data Portal: https://opendata.vancouver.ca/
-   R Documentation: https://cran.r-project.org/web/packages/tidyverse/
